---
title: "Validating the smktrans transition probabilities"
subtitle: "England"
author: "Duncan Gillespie, Sheffield Addictions Research Group"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    code_folding: hide
    df_print: kable
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = TRUE,
                      fig.width = 9, fig.height = 5.5, dpi = 150)
```

```{r params}
# Everything the report depends on, set once. The STS waves and the modelled
# years both come off val_years, so they cannot drift apart.
val_years <- 2019:2023
val_ages  <- 16:89
boot_B    <- 1000
boot_seed <- 20260716

source("00_validation_utils.R")
```

# What this report does and does not claim

There are two different things in here and they are not equally strong.

**Validation** is comparing our estimates against data that had no part in
producing them. There is one of those: quitting, against the Smoking Toolkit
Study. That is section 2.

**Verification** is checking that the pipeline does to its inputs what we think
it does. That is initiation, partly, and relapse entirely. It can tell us the
code is behaving. It cannot tell us the answer is right.

The reason relapse only gets verification is that the Smoking Toolkit Study is a
repeat cross-section. It never sees the same person twice, so it cannot observe
anyone relapsing, which is a transition between two points in time for one
person. There is no independent source we can put next to our relapse
probabilities. They rest on one paper, Hawkins et al. 2010, and the most we can
do is check that we have transcribed and transported it correctly.

Initiation sits in between. The STS cannot observe initiation either, for the
same reason, but it can observe the number of current smokers rising with age,
and that rise is initiation net of quitting. That is a real quantity and it is
one we estimate, so section 3 is a genuine comparison over the ages where
prevalence is still going up. It says nothing about `p_start` itself.

# Quitting against the Smoking Toolkit Study

We build the same quantity from the STS that our `p_quit` is: the probability
that someone who was a smoker at the start of the year is not one at the end.
The STS side reconstructs "smoker a year ago" by turning current ex-smokers back
into smokers if their most recent quit attempt began within the last year, then
takes one minus the ratio of the two survey-weighted proportions.

```{r quit-sts}
data_tk <- sts_read_england(years = val_years, ages = val_ages)

# How many ex-smokers fall out of the denominator because q632b8 is missing.
# They are treated as long-term quitters, which biases p_quit down.
frac_missing_q632b8 <- sts_check_q632b8(data_tk)

quit_sts <- sts_boot(dt = data_tk, fn = sts_quit_by_age, value_var = "p_quit",
                     domain_ages = val_ages, B = boot_B, seed = boot_seed)
```

```{r quit-model}
quit_model <- stapm_load("quit_data_England_uncertainty.rds")
quit_ref <- pop_weight_by_age(quit_model, "p_quit", years = val_years, ages = val_ages)
setnames(quit_ref, "value", "p_quit")

if (all(c("p_quit_lower", "p_quit_upper") %in% names(quit_model))) {
  quit_ref[, lower := pop_weight_by_age(quit_model, "p_quit_lower", val_years, val_ages)$value]
  quit_ref[, upper := pop_weight_by_age(quit_model, "p_quit_upper", val_years, val_ages)$value]
}
```

```{r quit-plot}
plot_ages <- 16:80
sts_plot   <- quit_sts[age %in% plot_ages & !is.na(est) & frac_undefined <= 0.05]
model_plot <- quit_ref[age %in% plot_ages]

ggplot() +
  geom_ribbon(data = model_plot, aes(age, ymin = lower, ymax = upper,
                                     fill = "smktrans"), alpha = 0.25) +
  geom_line(data = model_plot, aes(age, p_quit, colour = "smktrans"), linewidth = 1) +
  geom_errorbar(data = sts_plot, aes(age, ymin = lower, ymax = upper,
                                     colour = "Smoking Toolkit Study"), width = 0, alpha = 0.5) +
  geom_point(data = sts_plot, aes(age, est, colour = "Smoking Toolkit Study"), size = 1.5) +
  scale_colour_manual(values = c(smktrans = "#1f78b4", `Smoking Toolkit Study` = "#33a02c")) +
  scale_fill_manual(values = c(smktrans = "#1f78b4")) +
  coord_cartesian(ylim = c(0, 1)) +
  theme_minimal() + theme(legend.position = "bottom") +
  labs(x = "Age", y = "P(quit)", colour = NULL, fill = NULL,
       title = "Probability of quitting smoking, England",
       subtitle = sprintf("Bands and bars are 95%%. STS %s.",
                          paste(range(val_years), collapse = "-")))
```

```{r quit-numbers}
cmp <- merge(sts_plot[, .(age, sts = est, sts_lo = lower, sts_hi = upper)],
             model_plot[, .(age, model = p_quit)], by = "age")
cmp[, inside := model >= sts_lo & model <= sts_hi]

data.table(
  Measure = c("Ages compared",
              "smktrans inside the STS 95% interval",
              "Median difference (smktrans - STS)",
              "Mean absolute difference",
              "Correlation over age",
              "Ex-smokers with no usable q632b8"),
  Value = c(nrow(cmp),
            sprintf("%.0f%%", 100 * mean(cmp$inside)),
            sprintf("%+.4f", median(cmp$diff <- cmp$model - cmp$sts)),
            sprintf("%.4f", mean(abs(cmp$model - cmp$sts))),
            sprintf("%.3f", cor(cmp$model, cmp$sts)),
            sprintf("%.1f%%", 100 * frac_missing_q632b8))
)
```

## What to be careful about

The STS estimate is not assumption-free either. Three things carry through into
the comparison above.

Ex-smokers who did not answer q632b8 stay classified as ex-smokers, so they are
treated as long-term quitters and drop out of the denominator. If not answering
is related to how recently someone quit, this pushes the STS p_quit down. The
size of that group is in the table.

Waves are mapped to calendar years arithmetically, on the basis that wave 1 was
November 2006 and there has been one wave a month since. Only whole years are
used, so a year that is part-covered is dropped rather than contributing a
seasonally biased average.

The model side is averaged over sex, IMD and year using ONS populations, because
the STS figure is a population average and a plain mean over the ten subgroups
is not the same thing. The population file stops at `r max(fread(val_paths$pop_file)$year)`,
so for later years we hold that year's age, sex and IMD structure. That is an
assumption about who lives in England, not about smoking.

# Net initiation against the Smoking Toolkit Study

`calculate_net_initiation` produces

$$p_{start}^{net}(a) = \frac{n_{current}(a+1) - n_{current}(a)}{n_{never}(a) + n_{former}(a)}$$

and because never, current and former sum to a constant, the denominator is just
the people who are not currently smoking. Dividing through gives

$$p_{start}^{net}(a) = \frac{prev(a+1) - prev(a)}{1 - prev(a)}$$

which is exactly what we can compute from STS prevalence. Both sides are period
quantities: `calculate_net_initiation` runs a separate synthetic cohort for each
year through that year's probabilities, so it is an age profile for a year, not
a real birth cohort, and that is what a cross-section gives us too.

```{r netinit}
sts_ages_wide <- 16:35   # read wider than we plot so the difference at the top
                         # of the plotted range uses a real neighbour

data_tk_init <- sts_read_england(years = val_years, ages = sts_ages_wide)

net_sts <- sts_boot(dt = data_tk_init, fn = sts_net_init_by_age,
                    value_var = "p_start_net", domain_ages = sts_ages_wide,
                    B = boot_B, seed = boot_seed)

# Net initiation only means "initiation" while prevalence is still rising. Past
# that the numerator goes negative, the model clamps it to zero, and we would be
# comparing a clamp against noise. Take the range from the data.
prev_point  <- sts_net_init_by_age(data_tk_init, sts_ages_wide)
last_rising <- suppressWarnings(max(prev_point[!is.na(p_start_net) & p_start_net > 0]$age))
init_ages   <- min(sts_ages_wide):last_rising
```

Prevalence stops rising after age **`r last_rising`**, so the comparison runs
over ages `r min(init_ages)` to `r max(init_ages)`.

```{r netinit-model}
net_model <- stapm_load("net_init_data_England_uncertainty.rds")

# calculate_net_initiation only runs for 2011:2019, so it will not cover the
# whole STS window. Use the overlap and say so.
model_years <- intersect(val_years, unique(net_model$year))
net_ref <- pop_weight_by_age(net_model, "p_start_net", years = model_years, ages = init_ages)
setnames(net_ref, "value", "p_start_net")
```

```{r netinit-plot}
sts_i   <- net_sts[age %in% init_ages & !is.na(est)]
model_i <- net_ref[age %in% init_ages]

ggplot() +
  geom_ribbon(data = sts_i, aes(age, ymin = lower, ymax = upper,
                                fill = "Smoking Toolkit Study"), alpha = 0.2) +
  geom_point(data = sts_i, aes(age, est, colour = "Smoking Toolkit Study"), size = 1.5) +
  geom_line(data = model_i, aes(age, p_start_net, colour = "smktrans"), linewidth = 1) +
  scale_colour_manual(values = c(smktrans = "#1f78b4", `Smoking Toolkit Study` = "#33a02c")) +
  scale_fill_manual(values = c(`Smoking Toolkit Study` = "#33a02c")) +
  theme_minimal() + theme(legend.position = "bottom") +
  labs(x = "Age", y = "P(net initiation)", colour = NULL, fill = NULL,
       title = "Net smoking initiation, England",
       subtitle = sprintf("smktrans %s vs STS %s. Only where prevalence is rising.",
                          paste(range(model_years), collapse = "-"),
                          paste(range(val_years), collapse = "-")))
```

```{r netinit-numbers}
cmp_i <- merge(sts_i[, .(age, sts = est, sts_lo = lower, sts_hi = upper)],
               model_i[, .(age, model = p_start_net)], by = "age")
data.table(
  Measure = c("Ages compared", "smktrans inside the STS 95% interval",
              "Median difference", "Correlation over age", "Years compared"),
  Value = c(nrow(cmp_i),
            sprintf("%.0f%%", 100 * mean(cmp_i$model >= cmp_i$sts_lo & cmp_i$model <= cmp_i$sts_hi)),
            sprintf("%+.5f", median(cmp_i$model - cmp_i$sts)),
            sprintf("%.3f", cor(cmp_i$model, cmp_i$sts)),
            paste(range(model_years), collapse = "-"))
)
```

```{r netinit-caveat, results = "asis"}
if (!identical(sort(model_years), sort(val_years))) {
  cat(sprintf(paste0("**Note.** The STS window is %s but net initiation is only modelled ",
                     "for %s, so this comparison rests on the overlap: %s. Widening the ",
                     "years in `calculate_net_initiation` would give a fairer test.\n\n"),
              paste(range(val_years), collapse = "-"),
              paste(range(net_model$year), collapse = "-"),
              paste(model_years, collapse = ", ")))
}
```

The prevalence curve is smoothed over age before it is differenced, rather than
differenced and then repaired. Differencing a noisy curve gives a very noisy
derivative, and this is the same mistake that `p_dense` was making until July
2026: it built the initiation density by differencing a step function and then
tried to patch the resulting zeros inside `p_smooth`, which turned 0.1% of noise
on the input into a 51% change in the published probabilities.

# Relapse against the Hawkins inputs

This is verification and not validation. Hawkins et al. 2010 is our input, so
comparing our output back to it cannot tell us whether our relapse probabilities
are right. It tells us whether the pipeline is doing what we think.

The check that does the work is an invariant rather than a comparison.
`prep_relapse` computes a weighted average of Hawkins values over the demographic
profile, and a weighted average cannot fall outside the range of the things it
averages. So for every combination of time since quit, age, sex and IMD, our
output has to sit inside the minimum and maximum of Hawkins across the covariate
combinations. Historically that is a hard invariant and a failure is a bug. For
forecast years it is not, because `relapse_forecast` deliberately scales by the
fitted trend and may legitimately leave the range.

```{r relapse, results = "hold"}
# Sourced into its own environment rather than this one. The script has to be
# runnable on its own, so it sets its own parameters and names its own objects,
# and some of those names (cmp, prof) are ones this report uses too. Keeping it
# contained means the two cannot tread on each other. Its messages, warnings and
# printed output still come through.
source("23_verify_relapse_against_hawkins.R", local = new.env(), echo = FALSE)
```

## What this cannot check

The relapse module rests entirely on one paper and one cohort: 1,578 people in
the British Household Panel Survey between 1991 and 2006. Everything above
confirms we are using it consistently. None of it confirms that those estimates
describe England in the 2020s.

Three things are worth noting:

The baseline is calibrated rather than published. Hawkins reports odds ratios but
not the intercept of its model, so we solve for the baseline that makes the
Hawkins cohort reproduce the paper's own Table 2. That solve assumes the
covariates are independent of each other, which they are not, and assumes an age
distribution matching the mean and standard deviation implied by Table 1. Getting
that age distribution wrong is easy and it moves the answer: the correction is
1.07 done properly and 1.20 if the distribution is truncated carelessly. Once the
moments are matched the shape barely matters, and five different shapes all land
between 1.072 and 1.076. Before calibration the table gave the Hawkins cohort
14.28% relapsing at one year against the paper's 15.10%.

Relapse is set to exactly zero at ten years, with no uncertainty attached, and
the ABM cannot tell that apart from a finding. It is a convention. The paper's
own Kaplan-Meier curve is still declining past ten years, and the zero in Table 2
rests on no relapses out of 180 people in a single year.

Ages under 18 take the age-18 values, which are the highest in the series.



# Summary

```{r summary}
data.table(
  Component = c("Quitting", "Net initiation", "Initiation (p_start)", "Relapse"),
  `Independent data?` = c("Yes, Smoking Toolkit Study",
                          "Yes, Smoking Toolkit Study",
                          "No",
                          "No"),
  `What we can say` = c(
    "Compared directly against an independent survey",
    sprintf("Compared over ages %d-%d, where prevalence is still rising", min(init_ages), max(init_ages)),
    "Not identifiable from a repeat cross-section. Only the net figure above.",
    "Transcription and transport from Hawkins et al. verified. Level not independently checked."
  )
)
```

```{r sessioninfo}
data.table(
  Item = c("R version", "smktrans", "data.table", "Report run", "Bootstrap seed", "Bootstrap draws"),
  Value = c(R.version.string,
            as.character(packageVersion("smktrans")),
            as.character(packageVersion("data.table")),
            format(Sys.time(), "%Y-%m-%d %H:%M"),
            boot_seed, boot_B)
)
```
