---
title: "Validating the smktrans transition probabilities"
subtitle: "England"
author: "Duncan Gillespie, Sheffield Addictions Research Group"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    code_folding: hide
    df_print: kable
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 9, fig.height = 5.5, dpi = 150)
# Every path in this project is written relative to the project root, but knitr
# runs chunks from the Rmd's own directory by default. Pin it, so the knit works
# the same as sourcing the scripts from the root does, however render is called.
knitr::opts_knit$set(root.dir = normalizePath(".."))
```

```{r params}
# Everything the report depends on, set once. The STS waves and the modelled
# years both come off val_years, so they cannot drift apart.
# Validation years. These are chosen to sit entirely inside the model's
# ESTIMATION window, for two reasons. First, initiation is only estimated to
# 2017: the forecast jumps off at last_year - 1, so 2018 onwards in the
# published outputs is trend continuation, and comparing a forecast against a
# survey tests the forecast, not the estimation. Second, the STS switched to
# telephone interviewing in April 2020 and barely sampled 16-17 year olds for
# the next two years (cells of 6-29 people), so any window touching 2020-21 is
# soft at exactly the ages that matter for initiation. Waves in 2013-2017 are
# face-to-face with full sampling, and their 12-month lookback reaches into 2012
# at the earliest, which is still inside the estimation window. A comparison
# against 2019+ waves is a forecast check and belongs in a separately labelled
# section if we want it, not here.
val_years <- 2013:2017
val_ages  <- 16:89
boot_B    <- 1000
boot_seed <- 20260716

source("transition_probability_validation/00_validation_utils.R")
```

# What this report does and does not claim

There are two different things in here and they are not equally strong.

**Validation** is comparing our estimates against data that had no part in
producing them. Throughout, the comparison years sit inside the model's
estimation window (initiation is estimated to 2017; 2018 onwards is forecast),
and before the STS's 2020 switch to telephone interviewing. So what is being
tested is the estimation, on the years it was actually estimated, against a
survey operating normally. Forecast performance is a different question and is
not tested here; the one look we have taken at it, against the 2019-2023 waves,
disagrees at the youngest ages in the direction a trend fitted to pre-2019 data
has to disagree with a world where youth uptake turned up after 2020. That is a
statement about the forecast and the data feeding it, not about the estimation.

There are two validations: quitting, in section 2, and net initiation, in
section 3, both against the Smoking Toolkit Study.

**Verification** is checking that the pipeline does to its inputs what we think
it does. That is `p_start` itself, and relapse entirely. It can tell us the
code is behaving. It cannot tell us the answer is right.

The reason relapse only gets verification is that the Smoking Toolkit Study is a
repeat cross-section. It never sees the same person twice, so it cannot observe
anyone relapsing, which is a transition between two points in time for one
person. There is no independent source we can put next to our relapse
probabilities. They rest on one paper, Hawkins et al. 2010, and the most we can
do is check that we have transcribed and transported it correctly.

Initiation sits in between. The STS cannot observe initiation either, for the
same reason, but it can observe how the number of current smokers changes with
age, and that change is initiation net of quitting and relapse. That is a real
quantity and it is one we estimate, so section 3 is a genuine comparison - over
the whole age range, on both sides of zero. The age where the net flow changes
sign, where the cohort's smoking prevalence peaks, is itself part of what is
being compared. What this cannot do is separate the gross flows: it says
nothing about `p_start` on its own.

# Quitting against the Smoking Toolkit Study

We build the same quantity from the STS that our `p_quit` is: the probability
that someone who was a smoker at the start of the year is not one at the end.

The STS asks smoking status as `smokstat`, which comes with four answers:
smoking, stopped more than a year ago, stopped in the past year, and never
smoked. So the people who were smoking a year ago are the ones smoking now plus
the ones who stopped in the past year, and

$$p_{quit} = \frac{P(\text{stopped in past year})}{P(\text{smoking now}) + P(\text{stopped in past year})}$$

read straight off the survey-weighted proportions. The denominator is the
counterfactual population: everyone who was in a position to quit.

Two things about how it is lined up against the model.

smktrans indexes `p_quit` on age at the **start** of the year, so someone
observed at 40 who stopped in the past year is evidence about `p_quit(39)`, and
so is a smoker observed at 40. Both go in at age minus one. On these data that
choice moves the estimate by around 45%, so it is not a detail.

The years need lining up too, for the same sort of reason, and that is set out
under the caveats below.

The denominator we can build is not quite the one we want, and this is why the
plot starts at 25. Writing the states at the start of the year as S (smoking),
F (former) and V (never), what we can count is

$$|S| + |F \rightarrow S| + |V \rightarrow S|$$

when `p_quit` wants $|S|$ alone. Turning the recent quitters back on adds
$S \rightarrow F$ back where it belongs, but nothing removes the people who were
**not** smoking a year ago and are smoking now: relapsers, and at young ages
people who had never smoked and have just started. The STS has no variable for
when a current smoking spell began, so they cannot be found. The denominator is
too big and the STS estimate comes out too low.

A synthetic cohort run through our own 2019 probabilities puts a number on it.
From 26 upwards it is 1 to 5%, comfortably inside the bootstrap interval and not
worth worrying about, and nearly all of it is relapsers. Below 25 it is 23%, a
third of that is new smokers rather than relapsers, and at 18 it reaches 28%. A
comparison that is 28% out by construction is not a comparison, so the plot
starts where the arithmetic holds.

```{r quit-sts}
data_tk <- sts_read_england(years = val_years, ages = val_ages)
year_wts <- sts_model_year_weights(unique(data_tk$xwave))

quit_domain <- sts_domain(val_ages, by_sex = TRUE)
quit_sts <- sts_boot(dt = data_tk, fn = sts_quit_by_age, value_var = "p_quit",
                     domain = quit_domain, B = boot_B, seed = boot_seed)
```

```{r quit-model}
quit_model <- stapm_load("quit_data_England_uncertainty.rds")
quit_ref <- pop_weight_by_age(quit_model, "p_quit", years = year_wts$year, ages = val_ages,
                              by = c("sex", "age"), year_weights = year_wts)
setnames(quit_ref, "value", "p_quit")

if (all(c("p_quit_lower", "p_quit_upper") %in% names(quit_model))) {
  quit_ref[, lower := pop_weight_by_age(quit_model, "p_quit_lower", year_wts$year, val_ages,
                                        by = c("sex", "age"), year_weights = year_wts)$value]
  quit_ref[, upper := pop_weight_by_age(quit_model, "p_quit_upper", year_wts$year, val_ages,
                                        by = c("sex", "age"), year_weights = year_wts)$value]
}
```

```{r quit-plot}
plot_ages <- 25:80
sts_plot   <- quit_sts[age %in% plot_ages & !is.na(est) & frac_undefined <= 0.05]
model_plot <- quit_ref[age %in% plot_ages]

ggplot() +
  geom_ribbon(data = model_plot, aes(age, ymin = lower, ymax = upper,
                                     fill = "smktrans"), alpha = 0.25) +
  geom_line(data = model_plot, aes(age, p_quit, colour = "smktrans"), linewidth = 1) +
  geom_errorbar(data = sts_plot, aes(age, ymin = lower, ymax = upper,
                                     colour = "Smoking Toolkit Study"), width = 0, alpha = 0.5) +
  geom_point(data = sts_plot, aes(age, est, colour = "Smoking Toolkit Study"), size = 1.5) +
  facet_wrap(~ sex) +
  scale_colour_manual(values = c(smktrans = "#1f78b4", `Smoking Toolkit Study` = "#33a02c")) +
  scale_fill_manual(values = c(smktrans = "#1f78b4")) +
  coord_cartesian(ylim = c(0, 1)) +
  theme_minimal() + theme(legend.position = "bottom") +
  labs(x = "Age at the start of the year", y = "P(quit)", colour = NULL, fill = NULL,
       title = "Probability of quitting smoking, England",
       subtitle = sprintf("Bands and bars are 95%%. STS waves %s.",
                          paste(range(val_years), collapse = "-")))
```

```{r quit-years}
# Which model years the STS window actually covers.
year_wts[, .(`Model year` = year, `Exposure (wave-years)` = round(exposure, 2),
             Weight = sprintf("%.3f", weight))]
```

```{r quit-numbers}
cmp <- merge(sts_plot[, .(age, sex, sts = est, sts_lo = lower, sts_hi = upper)],
             model_plot[, .(age, sex, model = p_quit)], by = c("age", "sex"))
cmp[, `:=`(inside = model >= sts_lo & model <= sts_hi, diff = model - sts)]

cmp[, .(`Ages compared` = .N,
        `smktrans inside STS 95%` = sprintf("%.0f%%", 100 * mean(inside)),
        `Median difference` = sprintf("%+.4f", median(diff)),
        `Mean absolute difference` = sprintf("%.4f", mean(abs(diff))),
        `Correlation over age` = sprintf("%.3f", cor(model, sts))),
    by = .(Sex = sex)]
```

## What to be careful about

The STS estimate is not assumption-free either. Four things carry through into
the comparison above.

**The years the STS reports on are not the years its waves are fielded in.** It
asks whether someone stopped in the last twelve months, not whether they stopped
during a particular calendar year. So a wave fielded in March 2019 is reporting
on quits going back to March 2018, and nine of those twelve months are 2018.
Pooling all of 2019's waves observes five and a half years' worth of 2018 and six
and a half of 2019.

The model is therefore averaged over the years the window actually covers,
weighted by how much of each one the survey saw, rather than over the years the
waves are stamped with. The weights are in the table above.

The only wave selection that would land exactly on whole calendar years is
December waves alone, since a December interview looks back over that calendar
year and no further. That is genuinely exact and it throws away eleven twelfths
of the sample, which widens the intervals three and a half times and leaves a
plot that says nothing. The weighting is worth 0.9% here, under a fiftieth of the
confidence interval, so this is a correction that costs nothing and buys very
little. It is applied because it is free and because the next person to widen the
year range will get a larger number without knowing to look for it.

Waves are mapped to calendar years arithmetically, anchored on wave 150 being
March 2019, which puts wave 1 in October 2006. That checks out against the old
code: it filtered `xwave >= 148` and called it "2019-", and wave 148 is January
2019 under this mapping. Only whole years are used, so a part-covered year is
dropped rather than contributing a seasonally biased average.

The model side is averaged over sex, IMD and year using ONS populations, because
the STS figure is a population average and a plain mean over the ten subgroups
is not the same thing. The population file stops at `r max(fread(val_paths$pop_file)$year)`,
so for later years we hold that year's age, sex and IMD structure. That is an
assumption about who lives in England, not about smoking.

# Net initiation against the Smoking Toolkit Study

`calculate_net_initiation` produces

$$p_{start}^{net}(a) = \frac{n_{current}(a+1) - n_{current}(a)}{n_{never}(a) + n_{former}(a)}$$

and because never, current and former sum to a constant, the denominator is just
the people who are not currently smoking. Dividing through gives

$$p_{start}^{net}(a) = \frac{prev(a+1) - prev(a)}{1 - prev(a)}$$

which we can compute from STS prevalence two ways. A single cross-section
differences prevalence over age, but its gradient compares different birth
cohorts, and with initiation having fallen the older cohorts carry more smoking,
so it overstates net initiation at older ages - exactly where the model is being
judged. With five survey years we can instead follow the same cohort from one
year to the next, prev(a+1, t+1) against prev(a, t), which cancels the cohort
effect by construction. The cohort-followed series is the primary comparison;
the cross-section is plotted as open circles so the size of the cohort effect is
visible rather than asserted. The model side needs no matching change: its own
probabilities move slowly within the window, so its period profile and cohort
profile are close.

```{r netinit}
sts_ages_wide <- 16:35   # read wider than we plot so the difference at the top
                         # of the plotted range uses a real neighbour

data_tk_init <- sts_read_england(years = val_years, ages = sts_ages_wide)

net_domain <- sts_domain(sts_ages_wide)

# Primary: follow the same birth cohort across survey years, which cancels the
# cohort effect a single cross-section carries. Secondary: the cross-section,
# kept on the plot so the size of that effect is visible rather than asserted.
net_sts  <- sts_boot(dt = data_tk_init, fn = sts_net_init_cohort,
                     value_var = "p_start_net", domain = net_domain,
                     B = boot_B, seed = boot_seed)
net_cross <- sts_boot(dt = data_tk_init, fn = sts_net_init_by_age,
                      value_var = "p_start_net", domain = net_domain,
                      B = boot_B, seed = boot_seed)

# Both sides are now free to go negative. calculate_net_initiation used to clamp
# at zero, which meant cutting the comparison off at the age where prevalence
# stopped rising; it does not clamp any more, so the whole range is comparable.
init_ages <- sts_ages_wide
```

Net initiation goes negative past the age where smoking prevalence peaks, which
is a real feature of the age profile rather than something to hide, so the
comparison runs over the whole range and both sides are free to fall below zero.

```{r netinit-model}
net_model <- stapm_load("net_init_data_England_uncertainty.rds")

# calculate_net_initiation only runs for 2011:2019, so it will not cover the
# whole STS window. Use the overlap and say so.
model_years <- intersect(val_years, unique(net_model$year))
net_ref <- pop_weight_by_age(net_model, "p_start_net", years = model_years, ages = init_ages)
setnames(net_ref, "value", "p_start_net")
```

```{r netinit-plot}
sts_i   <- net_sts[age %in% init_ages & !is.na(est)]
model_i <- net_ref[age %in% init_ages]

ggplot() +
  geom_ribbon(data = sts_i, aes(age, ymin = lower, ymax = upper,
                                fill = "Smoking Toolkit Study"), alpha = 0.2) +
  geom_point(data = sts_i, aes(age, est, colour = "Smoking Toolkit Study"), size = 1.5) +
  geom_point(data = net_cross[age %in% init_ages & !is.na(est)],
             aes(age, est, colour = "Smoking Toolkit Study"), size = 1.5, shape = 1) +
  geom_line(data = model_i, aes(age, p_start_net, colour = "smktrans"), linewidth = 1) +
  geom_hline(yintercept = 0, linewidth = 0.3, colour = "grey50") +
  scale_colour_manual(values = c(smktrans = "#1f78b4", `Smoking Toolkit Study` = "#33a02c")) +
  scale_fill_manual(values = c(`Smoking Toolkit Study` = "#33a02c")) +
  theme_minimal() + theme(legend.position = "bottom") +
  labs(x = "Age", y = "P(net initiation)", colour = NULL, fill = NULL,
       title = "Net smoking initiation, England",
       subtitle = sprintf("smktrans %s vs STS %s. Only where prevalence is rising.",
                          paste(range(model_years), collapse = "-"),
                          paste(range(val_years), collapse = "-")))
```

```{r netinit-numbers}
cmp_i <- merge(sts_i[, .(age, sts = est, sts_lo = lower, sts_hi = upper)],
               model_i[, .(age, model = p_start_net)], by = "age")
data.table(
  Measure = c("Ages compared", "smktrans inside the STS 95% interval",
              "Median difference", "Correlation over age", "Years compared"),
  Value = c(nrow(cmp_i),
            sprintf("%.0f%%", 100 * mean(cmp_i$model >= cmp_i$sts_lo & cmp_i$model <= cmp_i$sts_hi)),
            sprintf("%+.5f", median(cmp_i$model - cmp_i$sts)),
            sprintf("%.3f", cor(cmp_i$model, cmp_i$sts)),
            paste(range(model_years), collapse = "-"))
)
```

```{r netinit-caveat, results = "asis"}
if (!identical(sort(model_years), sort(val_years))) {
  cat(sprintf(paste0("**Note.** The STS window is %s but net initiation is only modelled ",
                     "for %s, so this comparison rests on the overlap: %s. Widening the ",
                     "years in `calculate_net_initiation` would give a fairer test.\n\n"),
              paste(range(val_years), collapse = "-"),
              paste(range(net_model$year), collapse = "-"),
              paste(model_years, collapse = ", ")))
}
```

The prevalence curve is smoothed over age before it is differenced, rather than
differenced and then repaired. Differencing a noisy curve gives a very noisy
derivative, and this is the same mistake that `p_dense` was making until July
2026: it built the initiation density by differencing a step function and then
tried to patch the resulting zeros inside `p_smooth`, which turned 0.1% of noise
on the input into a 51% change in the published probabilities.

# Relapse against the Hawkins inputs

This is verification and not validation. Hawkins et al. 2010 is our input, so
comparing our output back to it cannot tell us whether our relapse probabilities
are right. It tells us whether the pipeline is doing what we think.

The check that does the work is an invariant rather than a comparison.
`prep_relapse` computes a weighted average of Hawkins values over the demographic
profile, and a weighted average cannot fall outside the range of the things it
averages. So for every combination of time since quit, age, sex and IMD, our
output has to sit inside the minimum and maximum of Hawkins across the covariate
combinations. Historically that is a hard invariant and a failure is a bug. For
forecast years it is not, because `relapse_forecast` deliberately scales by the
fitted trend and may legitimately leave the range.

The script ends with a figure: the Hawkins envelope itself, per time since
quit, over age, with our output drawn on top. The grey ribbon is the min to max
across the 160 covariate combinations and the dashed line their mean. The blue
jump-off line sitting inside the ribbon is the invariant holding; the gap
between it and the dashed mean is the demographic re-weighting; and anywhere
the red forecast line leaves the ribbon is the trend asserting a relapse
probability no combination of Hawkins covariates produces, in exactly the cells
the breach table counts. It is drawn for a single stratum - the one holding the
worst forecast breach - because averaging over sex and IMD would blur
cell-level breaches back inside a mixed envelope.

```{r relapse, results = "hold", message = TRUE}
# Sourced into its own environment rather than this one. The script has to be
# runnable on its own, so it sets its own parameters and names its own objects,
# and some of those names (cmp, prof) are ones this report uses too. Keeping it
# contained means the two cannot tread on each other. Its messages, warnings and
# printed output still come through.
source("transition_probability_validation/23_verify_relapse_against_hawkins.R",
       local = new.env(), echo = FALSE)
```

## What this cannot check

The relapse module rests entirely on one paper and one cohort: 1,578 people in
the British Household Panel Survey between 1991 and 2006. Everything above
confirms we are using it consistently. None of it confirms that those estimates
describe England in the 2020s.

Three things are worth noting:

The baseline is calibrated rather than published. Hawkins reports odds ratios but
not the intercept of its model, so we solve for the baseline that makes the
Hawkins cohort reproduce the paper's own Table 2. That solve assumes the
covariates are independent of each other, which they are not, and assumes an age
distribution matching the mean and standard deviation implied by Table 1. Getting
that age distribution wrong is easy and it moves the answer: the correction is
1.07 done properly and 1.20 if the distribution is truncated carelessly. Once the
moments are matched the shape barely matters, and five different shapes all land
between 1.072 and 1.076. Before calibration the table gave the Hawkins cohort
14.28% relapsing at one year against the paper's 15.10%.

Relapse is set to exactly zero at ten years, with no uncertainty attached, and
the ABM cannot tell that apart from a finding. It is a convention. The paper's
own Kaplan-Meier curve is still declining past ten years, and the zero in Table 2
rests on no relapses out of 180 people in a single year.

Ages under 18 take the age-18 values, which are the highest in the series.

## What would settle it

The British Household Panel Survey is archived at the UK Data Service.
Reconstructing Hawkins' cohort and refitting their model would give us the
intercept directly and remove every assumption in the calibration, along with the
covariate joint distribution we currently have to assume away. Given that the
whole relapse module rests on this one paper, that is probably worth someone's
time.

# Where this leaves things

```{r verdict}
# cmp_i is the section 3 comparison table; the numbers here are the same ones,
# so this paragraph cannot drift from the table above it.
v_cov  <- 100 * mean(cmp_i$model >= cmp_i$sts_lo & cmp_i$model <= cmp_i$sts_hi)
v_corr <- cor(cmp_i$model, cmp_i$sts)
v_med  <- median(cmp_i$model - cmp_i$sts)
```

The estimation stands up. Quitting sits inside the STS interval across ages 25
to 80 for both sexes. Net initiation, followed through cohorts, has the model
inside the STS 95% interval at `r sprintf("%.0f%%", v_cov)` of ages, a
correlation over age of `r sprintf("%.3f", v_corr)`, and a median difference of
`r sprintf("%+.5f", v_med)` - which is smaller than the smoothing bias we
measured in the cohort estimator itself on synthetic data (about 0.002). At that
point the ruler, not the model, is the binding constraint, and chasing the
difference further would mean fitting the ruler's error.

What this does not cover, deliberately: the forecast years, which are known to
miss the post-2020 turn in youth uptake because the survey feeding the trend
ends in 2018; and the level of relapse, which rests on Hawkins et al. alone.
Those are the two places any remaining worry should live. The STS decline in
prevalence at ages 31-34, visible past the end of the model line in section 3,
is close to a pure quit-versus-relapse quantity and would give relapse an
external check if the initiation pipeline were extended past age 30.

# Summary

```{r summary}
data.table(
  Component = c("Quitting", "Net initiation", "Initiation (p_start)", "Relapse"),
  `Independent data?` = c("Yes, Smoking Toolkit Study",
                          "Yes, Smoking Toolkit Study",
                          "No",
                          "No"),
  `What we can say` = c(
    "Compared directly against an independent survey",
    sprintf("Compared over ages %d-%d, on both sides of the prevalence peak", min(init_ages), max(init_ages)),
    "Not identifiable from a repeat cross-section. Only the net figure above.",
    "Transcription and transport from Hawkins et al. verified. Level not independently checked."
  )
)
```

```{r sessioninfo}
data.table(
  Item = c("R version", "smktrans", "data.table", "Report run", "Bootstrap seed", "Bootstrap draws"),
  Value = c(R.version.string,
            as.character(packageVersion("smktrans")),
            as.character(packageVersion("data.table")),
            format(Sys.time(), "%Y-%m-%d %H:%M"),
            boot_seed, boot_B)
)
```
